---
title-block-banner: true
title-block-banner-color: "#3F3F3F"
title: "Shifts in Values, Shifts in Democracy"
subtitle: "Identifying Correlations between World Values Survey Responses and Level of Electoral Democracy"
author:
  - name: Jason Gordon
    email: jagordon@uchicago.edu"
    url: https://github.com/jgsbd210945/30100_final
    affiliation: 
      - name: The University of Chicago
        city: Chicago
        state: IL
        url: https://www.uchicago.edu/
date: March 13, 2025
format: 
  pdf:
    keep-tex: true
    latex-engine: xelatex
bibliography: 30100_bibliography.bib
csl: chicago-fullnote-bibliography.csl
---

\newpage
\tableofcontents
\newpage

## Intoduction

Democratic backsliding has increasingly taken hold of the world, with the Varieties of Democracy (V-DEM) 2024 Democracy Report identifying 28 cases of backsliding in democracies, 13 of which having reverted to autocracies since the episode began [@nordDemocracyReport20242024]. As backsliding occurs within democracies with elected leaders, there is seemingly popular support *against* democracy in these countries — at least given the results of these elections. So, what shifts in an electorate's values are the most important to backsliding, and can we use machine learning techniques to predict the onset of backsliding?

In this paper, I plan to inspect if the wave of autocratization in democracies is the result of a shift in values against democracy, and which values are the most important for the shift to autocracy. As the number of cases of backsliding increases, it becomes increasingly important to understand why people would vote for actors that seek to challenge the core foundation of democratic institutions; with this knowledge, we may be able to further understand when cases of backsliding can happen.

First, I will be summarizing existing literature on democratic backsliding, identifying the tension in what causes backsliding. Then, I will conduct some exploratory data analysis and model fitting of my dataset, using Lasso to identify notable variables. Finally, I will discuss if these variables have importance in conducting future research.

## Existing Literature

Over the past twenty-five years, there has been considerable discussion about the causes of democratic backsliding - defined generally as the gradual breakdown of democratic institutions through nonviolent — and often, legal — means.[@haggardBackslidingDemocraticRegress2021; @levitskyHowDemocracies2018a] Most discussions claim polarization is the core idea behind why democracies denigrate, but little is known about the shifts in values to *why* it happens.

Haggard and Kaufman (2021a) discuss backsliding through the lens of dysfunction of a lack of trust in institutions, which permit autocrats to capture the executive.[@haggardAnatomyDemocraticBacksliding2021] These scholars point to an overall frustration with the system leading to polarization, as anti-system actors may be able to succeed in taking power if the population has no faith that the system will perform well with them. Other literature seems to correlate parts of this claim, with Cooley and Nexon (2020) arguing that these are akin to a "counter-order" movement, made in rejection to the liberal system of ordering.[@cooleyExitHegemonyUnraveling2020] Generally, these scholars agree that movements and politicians rejecting the structure of governance as it stands currently obtain success, with Haggard and Kaufman finding it is due to dissatisfaction in institutions.

However, Carothers and Press (2022) expand on this discussion, first limiting their discussion to the Global South and post-communist Europe, and delivering three separate categories of democratic backsliding, all of which create differing interpretations of the values behind backsliding. They state that backsliding can be caused by grievance-fueled illiberalism, agreeing with future literature that frustrations in institutions can drive backsliding; opportunistic authoritarianism, where autocrats act as political entrepreneurs to obtain power; and entrenched-interest revanchinism, where an interest group uses undemocratic means to reassert its claim. [@carothersUnderstandingRespondingGlobal2022] Carothers and Press's framework gives an idea for the potential variance in the viewpoints behind backsliding, claiming it is not as simple as people being mad about institutions failing them.

An additional perspective claims that backsliding isn't domestic in nature, but international. One such scholar, Anna Meyerrose (2020; 2021), finds that instead of backsliding being triggered by domestic parties, that international organizations (IOs) are to blame. She claims that because IOs focus on elections and an increase in executive power in their democracy promotion, in addition to removing some areas of policy for political parties to divide on, potential autocrats have an easier time gaining and maintaining power. [@meyerroseInternationalSourcesDemocratic2021; @meyerroseUnintendedConsequencesDemocracy2020] Deviating from other arguments, this branch of literature finds that while there may be domestic issues, structural problems make backsliding *much* more likely. As a result, backsliding may have nothing to do with values populations have.

I intend to put these differing claims to the test. While I expect that grievances with institutions play a not insignificant role in democratic backsliding, I do expect there to be variance in how much of a role it plays. Some states may backslide due to structural issues, or for electing a leader who at first may adhere to democratic norms, but then turn against the system later in their term.

## Data

```{r setup, include = FALSE}
knitr::opts_chunk$set(warnings = FALSE)
library(tidyverse)
library(vdemdata)
library(glmnet)
library(tidymodels)
library(xgboost)

load('Data/WVS_Cross-National_Wave_7_rData_v6_0.rdata')
wvs7 <- `WVS_Cross-National_Wave_7_v6_0` |> as.tibble()
```
I plan to utilize the World Values Survey's (WVS) seventh wave of data to account for the values I plan to inspect, pairing the averages of responses with the correlating electoral democracy scores found in the Varieties of Democracy's (V-DEM) data set.

While both are relatively large sources of data, combining the two results in just 62 observations; since WVS uses different questions between years, I cannot discuss it over time, leading to, unfortunately, a rather limited amount of data to analyze. Despite this, there are few other limitations between the two data sets. Both plot their results on a scale, meaning there are no significant outlying values, and they are both trusted sources of data used across the literature.

While I will focus on all questions in the WVS data, I plan on focusing on the electoral democracy and regime scores in the V-DEM data set. The electoral democracy index seeks to identify the structural components in a democracy; that is, are elections free and fair, and are the rights required for democratic rule generally followed? Alongside this, I will be collecting the overall regime type in the data, which puts all states on a ten-point scale from most to least democratic.

I will also create two additional variables, which I also aim to test to understand if a state is undergoing democratic backsliding: the difference in electoral democracy score, and a Boolean value for if a state is in a period of backsliding, defined below:

```{r chunk1}
workingvdem <- tibble(vdem) |>
  filter(year > 1975) |> # So I can do the lags correctly. This will be standardized later.
  group_by(country_text_id) |>
  arrange(year) |> # Should already be the case, but just in case it's not
  mutate(diff_polyarchy = v2x_polyarchy - lag(v2x_polyarchy)) |>
  # Backsliding boolean variable. I'm quantitatively defining it as having decreased in electoral democracy score by more than .005 and having decreased by at least 0.03 in the last two years.
  # I also filtered so that the regime has to be at least some semblance of an electoral autocracy (at least) so that hard autocracies getting more autocratic aren't included.
  mutate(backslided = (diff_polyarchy < -0.005) &
           (v2x_regime_amb > 2) &
           (lag(v2x_polyarchy, 2) - v2x_polyarchy > 0.03)) |>
  # Filling it out
  mutate(backslided = (diff_polyarchy < -0.001) &
           (backslided |
              ((lag(backslided) & lag(backslided, 2))|
                 lead(backslided) & lag(backslided) |
                 lead(backslided, 2) & lead(backslided)))) |>
  ungroup() |>
  filter(year > 2009) |>
  arrange(country_text_id) |>
  select(country_name, country_text_id, year, v2x_polyarchy, v2x_libdem, v2x_regime_amb, diff_polyarchy, backslided)
```

By defining a period of backsliding in an overall shift from democracy that does *not* have to be a sharp decline every year, I can create clear periods of backsliding, easily identifying cases.

In addition to creating these variables, I will be cleaning both data sets and attempting to standardize missing values if I cannot remove them, which will hopefully make my data as accurate as possible. This should also allow me to merge the V-DEM and WVS data without hassle.

```{r cleaningvdem, include = FALSE}
# 2022/2023 values (is NA for ones in backsliding. Let's make that True.)
workingvdem$backslided <- ifelse(is.na(workingvdem$backslided),
                                 TRUE,
                                 workingvdem$backslided)

isnum_lowna <- apply(workingvdem, 2, function(col){
  sum(is.na(col))
})
# The 1 is for South Sudan here.
na_rows <- apply(workingvdem, 1, function(row){sum(is.na(row))})
missing_rows <- na_rows != 0
workingvdem <- workingvdem[!missing_rows,]

wvs_test <- dplyr::select(wvs7, A_YEAR, B_COUNTRY_ALPHA, matches("^Q([0-9]+)")) |>
  rename(year = A_YEAR, country_text_id = B_COUNTRY_ALPHA)

isnum <- sapply(wvs_test, is.numeric) # Cols where we're working with #'s
# For those cols, remove all vals less than 0.
wvs_test[isnum] <- lapply(wvs_test[isnum], \(x) ifelse(x < 0, NA, x))

# By country
wvs_test <- wvs_test |>
  group_by(year, country_text_id) |>
  summarize(across(everything(), \(x) mean(x, na.rm = TRUE)))
isnum <- sapply(wvs_test, is.numeric) # Cols where we're working with #'s
# For those cols, remove all vals less than 0.
wvs_test[isnum] <- lapply(wvs_test[isnum], \(x) ifelse(is.na(x), -1, x))

merged <- merge(workingvdem, wvs_test, by = c("year", "country_text_id")) |>
  as_tibble()
merged$backslided <- factor(merged$backslided)
```
### Exploratory Analysis

To determine which features are the most important to a shift in electoral democracy score, I plan on using a Lasso model, testing the electoral democracy score to get the most important variables. While using the difference in this score may seem more intuitive, its low variance results in the model being unable to get the top features.

```{r lassosetup, echo = FALSE}
grid <- 10^seq(10, -2, length = 100)
y <- merged$v2x_polyarchy
z <- merged$diff_polyarchy |> scale()
merged_lasso <- dplyr::select(merged, matches("^Q([0-9]+)"))

# Lasso
glmnet(data.matrix(merged_lasso), y, alpha = 1, lambda = grid) |> plot()

set.seed(1)
cv_out <- cv.glmnet(data.matrix(merged_lasso), y, alpha = 1) 
plot(cv_out)
```

Given our cross-validation graph indicates our most important features will be somewhere between 10 and 15, we can then discuss which variables are the most important, as seen below.

```{r results}
top_features <- coef(cv_out, s = "lambda.1se")
top_features <- top_features[-1,]
top_features <- top_features[top_features != 0]
sort(abs(top_features), decreasing = TRUE) |> head(15)
```

Thus, the most impactful questions on the democracy score are as follows, alongside the graph of how important they are (with further away from 0 indicating increased important as more factors are added):
* Q211 *(Openness to attending a political demonstration)*
* Q239 *(Opinion on religious law governing a country)*
* Q38 *(Is it a child's duty to take care of a sick parent?)*
* Q212 *(Openness to joining unofficial strikes)*
* Q271 *(If the respondent is living with their parents)*
* Q225 *(Frequency of opposition candidates prevented from running)*
* Q73 *(Confidence in country's parliament)*
* Q3 *(Importance of leisure time)*
* Q292I *(Belief politicians are incompetent or ineffective)*
* Q182 *(Is homosexuality justifiable?)*
* Q262 *(Age)*
```{r lassoresults, echo = FALSE}
merged_select <- merged_lasso |> dplyr::select(names(top_features))
lasso_ms <- glmnet(data.matrix(merged_select), y, alpha = 1, lambda = grid) # y is still well-being
par(mar = c(5, 4, 4, 8), xpd = TRUE) # Need space for the legend
plot(lasso_ms)

# Making the legend
lasso_coefs <- coef(lasso_ms)
var_names <- rownames(lasso_coefs)[-1]
colors <- seq_len(length(var_names))
legend("right", legend = var_names, col = colors, lty = 1, cex = 0.8, inset = c(-0.25, 0))
```

Many of these reflect core foundations of democracy, but there are some that seem more cultural. We can see notable trends in political demonstrations highly correlating with overall level of electoral democracy; Q211 and Q212 are among the first variables given coefficients, and Q211 has the largest overall effect. This makes sense, as generally, an increased level to take public action for political reasons would increase the level of accountability leaders have to their people. 

Additionally, Wave 7 denotes variables like confidence in parliament, belief politicians are incompetent or ineffective, and frequency of opposition candidates being prevented from running. These all directly tie into whether electoral democracy is healthy, as if opposition candidates are not allowed to run, overall level of democracy decreases, and the last of these variables is one of the metrics V-DEM also measures to indicate level of electoral democracy.

Finally, the belief that the government and its actors may be ineffective make sense to discuss both electoral democracy and backsliding. As a population's trust of its government falls, it often polarizes and seeks anti-system candidates, resulting in an increased likelihood of a potential autocrat's election.

## Modeling

To test if I can predict electoral democracy levels and backsliding, I utilized an XGBoosted tree, as it allows me to use the most important variables found via Lasso to predict its results. Thus, I can focus on just these elements, creating more accurate predictions than other models.
```{r preprocessing, include = FALSE}
model_data <- dplyr::select(merged, country_name, v2x_polyarchy, diff_polyarchy, backslided, matches(var_names))

set.seed(1)
data_split <- initial_split(model_data, prop = 0.5)
train_data <- training(data_split)
test_data  <- testing(data_split)

rec <- recipe(v2x_polyarchy ~ ., data = train_data) |>
  step_select(all_numeric()) |>
  update_role(diff_polyarchy, new_role = "ID")

## Cross-validation object
cv_folds <- vfold_cv(train_data, v = 5, repeats = 1)

# xgboost
model_xgboost <- boost_tree(trees = 100, mtry=5, learn_rate = tune()) |> 
  set_mode("regression") |> 
  set_engine("xgboost")

xg_grid <- grid_regular(learn_rate(), levels = 5)

wf <- workflow() |>
  add_model(model_xgboost) |>
  add_recipe(rec)

model_res <- wf |>
  tune_grid(resamples = cv_folds,
            grid = xg_grid,
            control = control_grid(save_pred = TRUE))
```












## Conclusion

While the global shift to autocracy is bolstered by an overall frustration in democratic institutions, it is by no means the only factor. This research confirms the theoretical claim that backsliding is correlated with an increase in frustrations with democratic institutions, but also offers other potential variables like democratic activism and gender equity, which have potential for further theoretical analysis.

My initial hypothesis of frustration in institutions causing backsliding seems to hold, as trust in institutions is strongly correlated with government type. In particular, V115 (that of confidence in government during Wave 6 of the World Values Survey) was the variable most strongly correlated with government type for that set of data.

However, Wave 7's results posit an additional variable: that of democratic activism. Two of the most correlated variables are specifically about being open to political action. While further research is required to confirm its causality, this correlation implicates backsliding not only occurs due to frustration with governments but also as a result of a low will to protest for political purposes.

Thus, the level of democracy in a state - at least to some extent - is affected by the values of its population. If people are dissatisfied with the government and will rarely take political action, an autocratic regime is more likely. Using these correlations, we may be able to predict backsliding for future states, furthering our understanding of this shift from democracy in an ever-changing world.

## References

::: {#refs}
:::
